Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-21 doc/invt.tex
add pt_hist.gp

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$,
we have a quadratic bias potential
from a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magntiude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined by
Eq. \eqref{eq:maxFk}.
%
This method will be referred to as
the adaptive Gaussian ensemble method below.

The use of polynomial basis makes
the bias potential smooth and differentiable,
and hence suitable for MD applications.\cite{zhu2012}
%
However, since we only used two modes
for controlling the mean and variance,
the method only works
for narrow distributions.
%
To sample a wider range,
we can construct an expanded ensemble with
multiple Gaussian distributions
centered at different places via
either the replica exchange
(or parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
%
While the mutual exchange probability
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, \mathcal U(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$\mathcal U(\mathbf x)$ is the unbiased potential energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probability from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{(j)}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(j)}/Z^{(i)}$
from the following approximate formula
%
\begin{align*}
  \ln\frac{ Z^{(j)} }{ Z^{(i)} }
  \approx
  \ln\frac{ \sigma_j }{ \sigma_i }
   -u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.
\note{For implementation, we can further simplify
\begin{align*}
  &-u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
   \\
   &=
   (c_1^{(i)} + c_1^{(j)}) \, \Delta z^*
   +
   (c_2^{(i)} - c_2^{(j)}) \, \frac{{\Delta z^*}^2 - 1} {\sqrt 2}
   ,
\end{align*}
where $\Delta z^* = [z_c^{(j)} - z_c^{(i)}]/[\sigma_z^{(j)} + \sigma_z^{(i)}]$.
}

@@

\subsection{\label{sec:potts}
Adaptive Gaussian ensemble}

As a proof of principle,
we tested the adaptive Gaussian ensemble method
introduced in Sec. \ref{sec:bandpass}
on the two-dimensional $L\times L$ $10$-state
Potts model\cite{wu1982, newman, wang2001, wang2001pre}
along the energy space.
%
We used $L = 32$ here.
%
We used $26$ Gaussian distributions of width $L$,
and spread them over $[-1.7L^2, -0.9L^2]$.
%
For configuration sampling,
we used the Wolff algorithm\cite{wolff1989, newman}
at temperature $c_1/\sigma_z$
followed by a Metropolis acceptance step
to correct for the quadratic potential, $R_2$.
%
The acceptance ratio for configuration moves
was no less than 90\%.

@@

%\begin{figure}[h]
%\begin{center}
%  \makebox[\linewidth][c]{
%    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
%  }
%  \caption{
%    \label{fig:sinc}
%    Updating kernels of the bandpass updating schemes
%    compared to a Gaussian with a similar width.
%    %
%    The lines are a guide to the eyes.
%    Inset. Updating matrix in the non-periodic case.
%    %
%    \note{This figure was produced by \texttt{doc/fig/sinc.gp}.
%      The data were computed by the program \texttt{prog/predict},
%      see particularly the function \texttt{intq\_save()}
%      in \texttt{prog/intq.h} for panel (a),
%      and the functions \texttt{savewin()}
%      and \texttt{savewinmat()}
%      in \texttt{prog/invt.h} for panel (b),
%      The data were saved under the directory \texttt{data/sinc}.
%      See the \texttt{make} object \texttt{sincsingle}
%      in the \texttt{Makefile} there.
%    }%
%  }
%\end{center}
%\end{figure}

%In Fig. \ref{fig:sinc},
%we give examples of the updating kernels
%of the bandpass updating schemes,
%compared to a Gaussian
%with a similar width, $\sigma$,
%%
%\begin{equation}
%  \sigma
%  \approx
%  \frac
%  {
%    g \, n
%  }
%  {
%    \sqrt{ 2 \, \pi } \, 2 \, K
%  }
%  .
%\label{eq:sigma_equiv}
%\end{equation}
%%
%%where $g = 1$ for the periodic case
%%or $g = 2$ otherwise.
%%

>>>>

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$,
we have a quadratic bias potential
from a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magntiude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined by
Eq. \eqref{eq:maxFk}.
%
This method will be referred to as
the adaptive Gaussian umbrella sampling below.

The use of polynomial basis makes
the bias potential smooth and differentiable,
and hence suitable for MD applications.\cite{zhu2012}
%
However, since we only used two modes
for controlling the mean and variance,
the method only works
for narrow distributions.
%
To sample a wider range,
we can construct an expanded ensemble with
multiple Gaussian distributions
centered at different places via
either the replica exchange
(or parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
%
While the mutual exchange probability
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, \mathcal U(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$\mathcal U(\mathbf x)$ is the unbiased potential energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probability from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{(j)}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(j)}/Z^{(i)}$
from the following approximate formula
%
\begin{align*}
  \ln\frac{ Z^{(j)} }{ Z^{(i)} }
  \approx
  \ln\frac{ \sigma_j }{ \sigma_i }
   -u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.
%
In the $1/t$ scheme,
%
\note{For implementation, we can further simplify
\begin{align*}
  &-u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
   \\
   &=
   (c_1^{(i)} + c_1^{(j)}) \, \Delta z^*
   +
   (c_2^{(i)} - c_2^{(j)}) \, \frac{{\Delta z^*}^2 - 1} {\sqrt 2}
   ,
\end{align*}
where $\Delta z^* = [z_c^{(j)} - z_c^{(i)}]/[\sigma_z^{(j)} + \sigma_z^{(i)}]$.
%
However, the $c_2$ term does not appear to improve histogram flatness. Ideas?
}

@@

\subsection{\label{sec:potts}
Adaptive Gaussian umbrella sampling}

As a proof of principle,
we tested the method of adaptive Gaussian umbrella sampling
introduced in Sec. \ref{sec:bandpass}
on the two-dimensional $L\times L$ ten-state
Potts model\cite{wu1982, newman, wang2001, wang2001pre}
along the energy space.
%
We used $L = 32$ and $26$ Gaussian distributions of
the same width $\sigma_E = L$ over $[-1.7L^2, -0.9L^2]$
(the spacing between neighboring Gaussian centers and the width
were roughly the same).
%
For configuration sampling,
although one can use the Metropolis algorithm,
we found it more efficient to use
the Wolff algorithm\cite{wolff1989, newman}
at temperature $c_1/\sigma_E$
followed by a Metropolis acceptance step
to correct for the quadratic potential, $R_2$
(the acceptance ratio was no less than 92\%).
%
The results after $10^5 L^2$ steps are shown in
Fig. \ref{fig:pt_hist}.
%
The Gaussian histograms were evenly spaced with similar width as intended,
and the overall energy histogram was roughly flat,
as shown in \ref{fig:pt_hist}(a).
%
The parameter $c_1/\sigma_E$ roughly represents the local temperature,
and it demonstrated
the backbending behavior\cite{kim2006, *kim2007}
typical in a first-order phase transitions,
as shown in Fig. \ref{fig:pt_hist}(b).
%
The parameter $c_2$ represents the amount of ``pressure''
to restrain the distribution width,
and its value was maximal around the center critical region,
as shown in Fig. \ref{fig:pt_hist}(c).

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/pt_hist.pdf}
  }
  \caption{
    \label{fig:pt_hist}
    Application of the adaptive Gaussian umbrella sampling
    on the $32 \times 32$ ten state Potts model.
    %
    (a) Normalized histograms (solid lines) from adaptive Gaussian umbrella sampling
    matched the intended distributions.
    The overall histogram (dashed line) was roughly flat.
    %
    (b) The parameter, $c_1/\sigma_E$, roughly corresponding to
    the statistical temperature, $\beta(E)$,
    demonstrated the backbending behavior.
    %
    (c) The parameter $c_2$ for restraining the variance
    was maximal around the critical region.
  }
\end{center}
\end{figure}

@@
